Associations between Chlorophyll \emph{a} and various Microcystin-LR Health Advisory Concentrations

### 2015-10-01 manuscript/manuscript.Rmd
cleaned up refs, added tables:  Final draft prior to full read through

Over the last decade or so numerous events and legislative activities have raised the public awareness of harmful algal blooms [@jetoo2015toledo; @rinta2009lake; NEED HABHRCA REFS].  Recent, the US Environmental Protection Agency (USEPA) has released suggested microcystin-LR (one of the more common toxins) concentrations that would trigger health advisories [@mcelhiney2005detection; @zurawell2005hepatotoxic].  Additionally, the World Health Organization (WHO) has had proposed advisory levels for drinking water and a range of recreational risk levels [REF].   While these levels and associated advisories are likely to help mitigate the impacts from harmful algal blooms, they are not without complications.

@@

Chlorophyll *a* is a very commonly measured components of water quality that is also known to be associated with Microsystin-LR concentrations [@lee2000variation, @paerl2013harmful].  Additionally there are many rapid measurements for assessing chlorophyll *a* levels *in situ*.  For instance, there are small or hand held flourometers that provide reliable measurements [REFS].   Given these facts, it might be possible to identify chlorophyll *a* concentrations that would be associated with the various Microcystin-LR health advisory levels.  Identifying these associations would provide another reliable tool for water resource managers to use to help manage the threat to public health posed by cHABs and would be especially useful in the absence of microcystin-LR concentrations.  Thus, the goal of this paper is to utlize the National Lakes Assessment data and identify chlorophyll *a* concentrations that are associated with higher probabilities of exceeding several microcystin-LR health advisory concentrations [NLA REF].  So that others may repeat or adjust this analysis, the data, code, and this manuscript are freely available via [https://github.com/USAPE/microcystinchla](https://github.com/USAPE/microcystinchla).

@@

##Data
We used the 2007 National Lakes Assessment (NLA) water quality and microcystin-LR concentration data [REF]. These data represent a snapshot of water quality from the summer of 2007 and data on chlorophyll *a* and microcystin-LR concentrations are available for `r nrow(na.omit(nla_dat$CHLA,nla_dat$MCYST_TL_UGL))` lakes.

@@

A detailed discussion of CPA is beyond the scope of this paper, but see Paul et al. [-REF] and Hollister et al. [-REF] for details.  For this analysis, we used CPA to examine how the conditional probability of exceeding one of the health advisory changes as chlorophyll *a* increases in a lake.  The 95% confidence intervals were calculated from 1000 bootstrapped samples. To identify chlorophyll *a* concentrations of concern we used a 50% conditional probability of exceeding each health advisory level and extracted the minimum chlorophyll *a* concentration that was associated with the upper confidence level being 50% or greater.  As both microcystin-LR and chlorophyll *a* values were both highely skewed right, a log base 10 transformation was used.  Additional details of the specific implementation are available at https://github.com/USEPA/microcystinchla.

@@

Furthermore, the chlorophyll *a* cutoffs may be used to predict whether or not a lake exceeds the microcystin-LR health advisories.  Doing so allows us to compare the accuracy of the prediction as well as evaluate false negatives.  Total accuracy of the four cutoffs predicting microcystin-LR exceedances were `r total_accuracy$epa_child*100`% for the USEPA children's advisory, `r total_accuracy$who_drink*100`% for the WHO drinking water advisory,  `r total_accuracy$epa_adult*100`% for the USEPA adult advisory and, `r total_accuracy$who_rec*100`% for the WHO regreational advisory.  But Total Accuracy is one part of the prediction performace with which we are concerned.

When using the chlorophyll *a* cutoffs as an indicator of microcystin-LR exceedances, the error that should be avoided is predicting that no exceedance has occurred when in fact it has.  In other words, we would like to avoid Type II errors and minimize the proportion of false negatives [REF].  For the four chlorophyll *a* cut-offs we had a proportion of false negatives of `r fn_percent$epa_child*100`, `r fn_percent$who_drink*100`, `r fn_percent$epa_adult*100` and , `r fn_percent$who_rec*100 for the U.S EPA childrens drinking water, the WHO drinking water, the U.S. EPA Adult drinking water, and the WHO recreational advisories, respecitvely.  In each case we miss less than 10% of the lakes that are in fact exceeding the microcystin-LR advisory.  This is an acceptable rate, especially when no other tests are available for a given water body.

@@

>>>>

Over the last decade or so numerous events and legislative activities have raised the public awareness of harmful algal blooms [@jetoo2015toledo; @rinta2009lake; @HABHRCA2014].  Recently, the US Environmental Protection Agency (USEPA) has released suggested microcystin-LR (one of the more common toxins) concentrations that would trigger health advisories [@mcelhiney2005detection; @zurawell2005hepatotoxic; @usepa2015drinking].  Additionally, the World Health Organization (WHO) has had proposed advisory levels for drinking water and a range of recreational risk levels [@world2003cyanobacterial; @chorus1999toxic].   While these levels and associated advisories are likely to help mitigate the impacts from harmful algal blooms, they are not without complications.

@@

Chlorophyll *a* is a very commonly measured components of water quality that is also known to be associated with Microsystin-LR concentrations [@lee2000variation; @paerl2013harmful; heisler2008eutrophication].  Additionally there are many rapid measurements for assessing chlorophyll *a* levels *in situ*.  For instance, there are small or hand held flourometers that measure chlorohpyll *a*.   Given these facts, it might be possible to identify chlorophyll *a* concentrations that would be associated with the various Microcystin-LR health advisory levels.  Identifying these associations would provide another reliable tool for water resource managers to use to help manage the threat to public health posed by cHABs and would be especially useful in the absence of microcystin-LR concentrations.  Thus, the goal of this paper is to utlize the National Lakes Assessment data and identify chlorophyll *a* concentrations that are associated with higher probabilities of exceeding several microcystin-LR health advisory concentrations [@usepa2009national].  So that others may repeat or adjust this analysis, the data, code, and this manuscript are freely available via [https://github.com/USAPE/microcystinchla](https://github.com/USAPE/microcystinchla).

@@

##Data
We used the 2007 National Lakes Assessment (NLA) water quality and microcystin-LR concentration data [@usepa2009national]. These data represent a snapshot of water quality from the summer of 2007 and data on chlorophyll *a* and microcystin-LR concentrations are available for `r nrow(na.omit(nla_dat$CHLA,nla_dat$MCYST_TL_UGL))` lakes.

@@

A detailed discussion of CPA is beyond the scope of this paper, but see Paul et al. [-@paul2005development] and Hollister et al. [-@hollister2008cprob] for details.  For this analysis, we used CPA to examine how the conditional probability of exceeding one of the health advisory changes as chlorophyll *a* increases in a lake.  The 95% confidence intervals were calculated from 1000 bootstrapped samples. To identify chlorophyll *a* concentrations of concern we used a 50% conditional probability of exceeding each health advisory level and extracted the minimum chlorophyll *a* concentration that was associated with the upper confidence level being 50% or greater.  As both microcystin-LR and chlorophyll *a* values were both highely skewed right, a log base 10 transformation was used.  Additional details of the specific implementation are available at https://github.com/USEPA/microcystinchla.

@@

Furthermore, the chlorophyll *a* cutoffs may be used to predict whether or not a lake exceeds the microcystin-LR health advisories.  Doing so allows us to compare the accuracy of the prediction as well as evaluate false negatives.  Total accuracy of the four cutoffs predicting microcystin-LR exceedances were `r total_accuracy$epa_child*100`% for the USEPA children's advisory, `r total_accuracy$who_drink*100`% for the WHO drinking water advisory,  `r total_accuracy$epa_adult*100`% for the USEPA adult advisory and, `r total_accuracy$who_rec*100`% for the WHO regreational advisory (Tables \ref{tab:child_conmat_table}, \ref{tab:who_drink_conmat_table}, \ref{tab:adult_conmat_table}, & \ref{tab:who_rec_conmat_table}).  But Total Accuracy is one part of the prediction performace with which we are concerned.

When using the chlorophyll *a* cutoffs as an indicator of microcystin-LR exceedances, the error that should be avoided is predicting that no exceedance has occurred when in fact it has.  In other words, we would like to avoid Type II errors and minimize the proportion of false negatives.  For the four chlorophyll *a* cut-offs we had a proportion of false negatives of `r fn_percent$epa_child*100`%, `r fn_percent$who_drink*100`%, `r fn_percent$epa_adult*100`% and , `r fn_percent$who_rec*100`% for the U.S EPA childrens drinking water, the WHO drinking water, the U.S. EPA Adult drinking water, and the WHO recreational advisories, respecitvely.  In each case we miss less than 10% of the lakes that are in fact exceeding the microcystin-LR advisory.  This is an acceptable rate, especially when no other tests are available for a given water body.

@@

```{r child_conmat_table, results='asis', echo=FALSE, tab.cap="Confusion matrix comparing chlorophyll \\textit{a} predicted exceedences (rows) versus real exceedances (columns) for the U.S. EPA childrens drinking water advisory. \\label{tab:child_conmat_table}"}
knitr::kable(child_conmat)
```

```{r who_drink_conmat_table, results='asis', echo=FALSE, tab.cap="Confusion matrix comparing chlorophyll \\textit{a} predicted exceedences (rows) versus real exceedances (columns) for the WHO drinking water advisory. \\label{tab:who_drink_conmat_table}"}
knitr::kable(who_drink_conmat)
```

```{r adult_conmat_table, results='asis', echo=FALSE, tab.cap="Confusion matrix comparing chlorophyll \\textit{a} predicted exceedences (rows) versus real exceedances (columns) for the U.S. EPA adult drinking water advisory. \\label{tab:adult_conmat_table}"}
knitr::kable(adult_conmat)
```

```{r who_rec_conmat_table, results='asis', echo=FALSE, tab.cap="Confusion matrix comparing chlorophyll \\textit{a} predicted exceedences (rows) versus real exceedances (columns) for the WHO recreational water advisory. \\label{tab:who_rec_conmat_table}"}
knitr::kable(who_rec_conmat)
```

### 2015-10-01 manuscript/manuscript.tex
cleaned up refs, added tables:  Final draft prior to full read through

Over the last decade or so numerous events and legislative activities
have raised the public awareness of harmful algal blooms {[}Jetoo et al.
(2015); Rinta-Kanto et al. (2009); NEED HABHRCA REFS{]}. Recent, the US
Environmental Protection Agency (USEPA) has released suggested
microcystin-LR (one of the more common toxins) concentrations that would
trigger health advisories (McElhiney and Lawton 2005, Zurawell et al.
2005). Additionally, the World Health Organization (WHO) has had
proposed advisory levels for drinking water and a range of recreational
risk levels {[}REF{]}. While these levels and associated advisories are
likely to help mitigate the impacts from harmful algal blooms, they are
not without complications.

@@

Chlorophyll \emph{a} is a very commonly measured components of water
quality that is also known to be associated with Microsystin-LR
concentrations (Lee et al. 2000, Paerl and Otten (2013)). Additionally
there are many rapid measurements for assessing chlorophyll \emph{a}
levels \emph{in situ}. For instance, there are small or hand held
flourometers that provide reliable measurements {[}REFS{]}. Given these
facts, it might be possible to identify chlorophyll \emph{a}
concentrations that would be associated with the various Microcystin-LR
health advisory levels. Identifying these associations would provide
another reliable tool for water resource managers to use to help manage
the threat to public health posed by cHABs and would be especially
useful in the absence of microcystin-LR concentrations. Thus, the goal
of this paper is to utlize the National Lakes Assessment data and
identify chlorophyll \emph{a} concentrations that are associated with
higher probabilities of exceeding several microcystin-LR health advisory
concentrations {[}NLA REF{]}. So that others may repeat or adjust this
analysis, the data, code, and this manuscript are freely available via
\href{https://github.com/USAPE/microcystinchla}{\url{https://github.com/USAPE/microcystinchla}}.

@@

We used the 2007 National Lakes Assessment (NLA) water quality and
microcystin-LR concentration data {[}REF{]}. These data represent a
snapshot of water quality from the summer of 2007 and data on
chlorophyll \emph{a} and microcystin-LR concentrations are available for
lakes.

@@

A detailed discussion of CPA is beyond the scope of this paper, but see
Paul et al. {[}-REF{]} and Hollister et al. {[}-REF{]} for details. For
this analysis, we used CPA to examine how the conditional probability of
exceeding one of the health advisory changes as chlorophyll \emph{a}
increases in a lake. The 95\% confidence intervals were calculated from
1000 bootstrapped samples. To identify chlorophyll \emph{a}
concentrations of concern we used a 50\% conditional probability of
exceeding each health advisory level and extracted the minimum
chlorophyll \emph{a} concentration that was associated with the upper
confidence level being 50\% or greater. As both microcystin-LR and
chlorophyll \emph{a} values were both highely skewed right, a log base
10 transformation was used. Additional details of the specific
implementation are available at
\url{https://github.com/USEPA/microcystinchla}.

@@

Furthermore, the chlorophyll \emph{a} cutoffs may be used to predict
whether or not a lake exceeds the microcystin-LR health advisories.
Doing so allows us to compare the accuracy of the prediction as well as
evaluate false negatives. Total accuracy of the four cutoffs predicting
microcystin-LR exceedances were 74\% for the USEPA children's advisory,
86\% for the WHO drinking water advisory, 90\% for the USEPA adult
advisory and, 91\% for the WHO regreational advisory. But Total Accuracy
is one part of the prediction performace with which we are concerned.

When using the chlorophyll \emph{a} cutoffs as an indicator of
microcystin-LR exceedances, the error that should be avoided is
predicting that no exceedance has occurred when in fact it has. In other
words, we would like to avoid Type II errors and minimize the proportion
of false negatives {[}REF{]}. For the four chlorophyll \emph{a} cut-offs
we had a proportion of false negatives of 9, 7, 6 and , `r
fn\_percent\$who\_rec*100 for the U.S EPA childrens drinking water, the
WHO drinking water, the U.S. EPA Adult drinking water, and the WHO
recreational advisories, respecitvely. In each case we miss less than
10\% of the lakes that are in fact exceeding the microcystin-LR
advisory. This is an acceptable rate, especially when no other tests are
available for a given water body.

@@

@@

@@

@@

>>>>

Over the last decade or so numerous events and legislative activities
have raised the public awareness of harmful algal blooms (Rinta-Kanto et
al. 2009, HABHRCA 2014, Jetoo et al. 2015). Recently, the US
Environmental Protection Agency (USEPA) has released suggested
microcystin-LR (one of the more common toxins) concentrations that would
trigger health advisories (McElhiney and Lawton 2005, Zurawell et al.
2005, USEPA 2015). Additionally, the World Health Organization (WHO) has
had proposed advisory levels for drinking water and a range of
recreational risk levels (Chorus and Bartram 1999, (WHO) and others
2003). While these levels and associated advisories are likely to help
mitigate the impacts from harmful algal blooms, they are not without
complications.

@@

Chlorophyll \emph{a} is a very commonly measured components of water
quality that is also known to be associated with Microsystin-LR
concentrations {[}Lee et al. (2000); Paerl and Otten (2013);
heisler2008eutrophication{]}. Additionally there are many rapid
measurements for assessing chlorophyll \emph{a} levels \emph{in situ}.
For instance, there are small or hand held flourometers that measure
chlorohpyll \emph{a}. Given these facts, it might be possible to
identify chlorophyll \emph{a} concentrations that would be associated
with the various Microcystin-LR health advisory levels. Identifying
these associations would provide another reliable tool for water
resource managers to use to help manage the threat to public health
posed by cHABs and would be especially useful in the absence of
microcystin-LR concentrations. Thus, the goal of this paper is to utlize
the National Lakes Assessment data and identify chlorophyll \emph{a}
concentrations that are associated with higher probabilities of
exceeding several microcystin-LR health advisory concentrations (USEPA
2009). So that others may repeat or adjust this analysis, the data,
code, and this manuscript are freely available via
\href{https://github.com/USAPE/microcystinchla}{\url{https://github.com/USAPE/microcystinchla}}.

@@

We used the 2007 National Lakes Assessment (NLA) water quality and
microcystin-LR concentration data (USEPA 2009). These data represent a
snapshot of water quality from the summer of 2007 and data on
chlorophyll \emph{a} and microcystin-LR concentrations are available for
lakes.

@@

A detailed discussion of CPA is beyond the scope of this paper, but see
Paul et al. (2005) and Hollister et al. (2008) for details. For this
analysis, we used CPA to examine how the conditional probability of
exceeding one of the health advisory changes as chlorophyll \emph{a}
increases in a lake. The 95\% confidence intervals were calculated from
1000 bootstrapped samples. To identify chlorophyll \emph{a}
concentrations of concern we used a 50\% conditional probability of
exceeding each health advisory level and extracted the minimum
chlorophyll \emph{a} concentration that was associated with the upper
confidence level being 50\% or greater. As both microcystin-LR and
chlorophyll \emph{a} values were both highely skewed right, a log base
10 transformation was used. Additional details of the specific
implementation are available at
\url{https://github.com/USEPA/microcystinchla}.

@@

Furthermore, the chlorophyll \emph{a} cutoffs may be used to predict
whether or not a lake exceeds the microcystin-LR health advisories.
Doing so allows us to compare the accuracy of the prediction as well as
evaluate false negatives. Total accuracy of the four cutoffs predicting
microcystin-LR exceedances were 74\% for the USEPA children's advisory,
86\% for the WHO drinking water advisory, 90\% for the USEPA adult
advisory and, 91\% for the WHO regreational advisory (Tables
\ref{tab:child_conmat_table}, \ref{tab:who_drink_conmat_table},
\ref{tab:adult_conmat_table}, \& \ref{tab:who_rec_conmat_table}). But
Total Accuracy is one part of the prediction performace with which we
are concerned.

When using the chlorophyll \emph{a} cutoffs as an indicator of
microcystin-LR exceedances, the error that should be avoided is
predicting that no exceedance has occurred when in fact it has. In other
words, we would like to avoid Type II errors and minimize the proportion
of false negatives. For the four chlorophyll \emph{a} cut-offs we had a
proportion of false negatives of 9\%, 7\%, 6\% and , 5\% for the U.S EPA
childrens drinking water, the WHO drinking water, the U.S. EPA Adult
drinking water, and the WHO recreational advisories, respecitvely. In
each case we miss less than 10\% of the lakes that are in fact exceeding
the microcystin-LR advisory. This is an acceptable rate, especially when
no other tests are available for a given water body.

@@

\begin{longtable}[c]{@{}lrr@{}}
\caption{Confusion matrix comparing chlorophyll \textit{a} predicted
exceedences (rows) versus real exceedances (columns) for the U.S. EPA
childrens drinking water advisory.
\label{tab:child_conmat_table}}\tabularnewline
\toprule
& FALSE & TRUE\tabularnewline
\midrule
\endfirsthead
\toprule
& FALSE & TRUE\tabularnewline
\midrule
\endhead
FALSE & 638 & 94\tabularnewline
TRUE & 173 & 123\tabularnewline
\bottomrule
\end{longtable}

\begin{longtable}[c]{@{}lrr@{}}
\caption{Confusion matrix comparing chlorophyll \textit{a} predicted
exceedences (rows) versus real exceedances (columns) for the WHO
drinking water advisory.
\label{tab:who_drink_conmat_table}}\tabularnewline
\toprule
& FALSE & TRUE\tabularnewline
\midrule
\endfirsthead
\toprule
& FALSE & TRUE\tabularnewline
\midrule
\endhead
FALSE & 838 & 76\tabularnewline
TRUE & 69 & 45\tabularnewline
\bottomrule
\end{longtable}

\begin{longtable}[c]{@{}lrr@{}}
\caption{Confusion matrix comparing chlorophyll \textit{a} predicted
exceedences (rows) versus real exceedances (columns) for the U.S. EPA
adult drinking water advisory.
\label{tab:adult_conmat_table}}\tabularnewline
\toprule
& FALSE & TRUE\tabularnewline
\midrule
\endfirsthead
\toprule
& FALSE & TRUE\tabularnewline
\midrule
\endhead
FALSE & 887 & 57\tabularnewline
TRUE & 50 & 34\tabularnewline
\bottomrule
\end{longtable}

\begin{longtable}[c]{@{}lrr@{}}
\caption{Confusion matrix comparing chlorophyll \textit{a} predicted
exceedences (rows) versus real exceedances (columns) for the WHO
recreational water advisory.
\label{tab:who_rec_conmat_table}}\tabularnewline
\toprule
& FALSE & TRUE\tabularnewline
\midrule
\endfirsthead
\toprule
& FALSE & TRUE\tabularnewline
\midrule
\endhead
FALSE & 910 & 52\tabularnewline
TRUE & 43 & 23\tabularnewline
\bottomrule
\end{longtable}

@@

Chorus, E. I., and J. Bartram. 1999. Toxic cyanobacteria in water: A
guide to their public health consequences, monitoring and management.

HABHRCA. 2014. Harmful algal bloom and hypoxia research and control
amendments act of 2014.

Hollister, J. W., H. A. Walker, and J. F. Paul. 2008. CProb: A
computational tool for conducting conditional probability analysis.
Journal of environmental quality 37:2392--2396.

@@

Paul, J. F., and M. E. McDonald. 2005. Development of empirical,
geographically specific water quality criteria: A conditional
probability analysis approach 41:1211--1223.

@@

USEPA. 2009. National lakes assessment: A collaborative survey of the
nation's lakes. ePA 841-r-09-001. Office of Water; Office of Research;
Development, US Environmental Protection Agency Washington, DC.

USEPA. 2015. Drinking water health advisory for the cyanobacterial
microcystin toxins. ePA-820R15100. Office of Water, US Environmental
Protection Agency Washington, DC.

(WHO), W. H. O., and others. 2003. Cyanobacterial toxins: Microcystin-lR
in drinking-water. background document for development of wHO guidelines
for drinking-water quality. geneva, switzerland. World Health
Organization, 2nd ed. Geneva.
